---
title: "Week 3 Lecture notes"
author: "Lynnsha Jing Huang"
date: "Monday, August 17, 2015"
output: html_document
runtime: shiny
---
### Loop Functions###
#### Looping on the Command Line ####

Writing for, while loops is useful when programming but not particularly easy when working interactively on the command line. There are some functions which implement looping to make life easier.

- **lapply**: Loop over a list and evaluate a function on each element
- **sapply**: Same as **lapply** but try to simplify the result
- **tapply** Apply a function over the margins of an array
- **mapply** Multivariate version of **lapply**

An auxilliary **split** is also useful, particularly in conjunction with **lapply**.

#### lapply ####
**lapply** takes three arguments: (1) a list **x**; (2) a function(or the nameof a function) **FUN**; (3) other arguments via its...argument. If **x** is not a list, it will be coerced to a list using **as.list**.

```
lapply
```
```
## function (X, FUN, ...)
## {
## FUN <- match.fun(FUN)
## if (!is.vector(X) || is.object(X))
## X <- as.list(X)
## .Internal(lapply(X, FUN))
## }
## <bytecode: 0x7ff7a1951c00>
## <environment: namespace:base>

```
The actual looping is done internally in C code.

**lapply** always returns a list, regardless of the class of the input.

```{R}
x <- list(a = 1:5, b = rnorm(10))
lapply(x, mean)
```
```{R}
x <- list(a = 1:4, b = rnorm(10), c = rnorm(20, 1), d = rnorm(100, 5))
lapply(x, mean)
```
```{R}
x <- 1:4
lapply(x, runif)

```

```{R]}
x<-1:4
lapply(x,runif, min=0, max=10)

```
**lapply** and friends make heavy use of *anonymous* functions
```{R}
x<-list(a=matrix(1:4, 2,2), b=matrix(1:6, 3,2))
x
```
An anonymous function for extracting the first column of each matrix.
```{R}
lapply(x, function(elt) elt[,1])
```
#### sapply ####
**sapply** will try to simplify the result of **lapply** if possible.

- If the result is a list where every element is length 1, then a vector is returned
- If the result is a list where every element is a vector of the same length(>1), a matrix is returned.
- If it can't figure things out, a list is returned.

```{R}
x <- list(a = 1:4, b = rnorm(10), c = rnorm(20, 1), d = rnorm(100, 5))
lapply(x, mean)
```
```{R}
sapply(x, mean)
mean(x)

```
#### apply ####
**apply** is used to evaluate a function(often an anonymous one) over the margins of an array.

- It is most often used to apply a function to the rows or columns of a matrix
- It can be used with general arrary, e.g. taking the average of an array of matrices
- It is not really faster than writing a loop, but it works in one line!

```
str(apply)
function(x, MARGIN, FUN, ...)
```
- **x** is an array
- **MARGIN** is an integer vector indicating which margins should be "retained- "FUN" is a function to be applied
-... is for other arguments to be passed to **FUN**

```{R}
x<-matrix(rnorm(200), 20, 10)
apply(x, 2, mean)
apply(x,1, sum)
```
#### col/row sums and means ####
For sums and means of matrix dimensions, we have some shortcuts.

- **rowSums** = **apply(x, 1, sum)**
- **rowMeans** = **apply(x, 1, mean)**
- **colSums** = **apply(x, 2, mean)**
The shortcut functions are *much* faster, but you won't notice unless you're using a large matrix.

#### Other Ways to Apply ####
Quantiles of the rows of a matrix

```{R}
x<-matrix(rnorm(200), 20, 10)
apply(x,1, quantile, probs=c(0.25, 0.75))

```
Average matrix in an array
```{R}
a<-array(rnorm(2*2*10), c(2,2,10))
apply(a, c(1,2), mean)

rowMeans(a, dims =2)
```
#### mapply ####
**mapply** is a multivariate apply of sorts which applies a function in parallel over a set of arguments.

```
str(mapply)
function(FUN, ..., MoreArgs = NULL, SIMPLIFY =TRUE,
          USE.NAMES= TRUE)
```
- **FUN** is a function to apply
- ...contains arguments to apply over
- **MoreArgs** is a list of other arguments to **FUN**
- **SIMPLIFY ** indicates whether the result should be simplified.

The following is tedious to type
**list(rep(1, 4), rep(2,3), rep(3, 2), rep(4, 1))**

Instead we can do
```{R}
mapply(rep, 1:4, 4:1)
```
#### Vectorizing a Function ####
```{R}
noise <-function(n, mean, sd) {
rnorm(n, mean, sd)
}
noise(5, 1, 2)
noise(1:5, 1:5, 2)
```
#### Instant Vectorization ####
Which is the same as 

```{R}
list(noise(1,1,2), noise(2,2,2), 
     noise(3,3,2), noise(4,4,2), 
     noise(5,5,2))
```
#### tapply ####
**tapply** is used to apply a function over subsets of a vector. I don't know why it's called **tapply**.

```
str(tapply)
function(x, INDEX, FUN=NULL,...,simplify=TRUE)
```
- **x** is a vector
- **INDEX** is a factor or a list of factors(or else they are coerced to factors)
- **FUN** is a function to be applied
- ... contains other arguments to be passed **FUN**
- **simplify**, should we simplify the result?

Take group means.
```{R}
x<- c(rnorm(10), runif(10), rnorm(10,1))
f<- gl(3, 10)
f
tapply(x,f,mean)
```
Take group means without simplification

```{R}
tapply(x, f, mean, simplify=FALSE)
```
#### tapply ####
Find group ranges.

```{r}
tapply(x, f, range)
```

#### split ####
split takes a vector or other objects and splits it into groups determined by a factor or list of factors. 
```
str(split)
function (x,f,drop=FALSE, ...)
```
-**x** is a vector(or list) or data frame
-**f** is a factor(or coerced to one)or a list of factors
-**drop** indicates whether empty factors lecels should be dropped

```{r}
x<- c(rnorm(10), runif(10), rnorm(10,1))
f <-gl(3, 10)
split(x, f)
```
A common idiom is **split** followed by an **lapply**
```{r}
lapply(split(x,f),mean)
```
#### Splitting a Data Frame ####

```{r}
library(datasets)
head(airquality)
```
```{r}
s <- split(airquality, airquality$Month)
lapply(s, function(x) colMeans(x[, c("Ozone", "Solar.R", "Wind")]))
```
```{r}
 sapply(s, function(x) colMeans(x[, c("Ozone", "Solar.R", "Wind")]))

 sapply(s, function(x) colMeans(x[, c("Ozone", "Solar.R", "Wind")],
 na.rm = TRUE))
```
#### Splitting on More than One Level ####
```{r}
x<-rnorm(10)
f1<-gl(2,5)
f2<-gl(5,2)
f1
f2
interaction(f1,f2)
```
#### Splitting on More than One Level ####
Interactions can create empty levels.
```{r}
str(split(x,list(f1, f2)))

```
Empty levels can be dropped
```{r}
str(split(x,list(f1,f2),drop=TRUE))
```
### Debugging ###
#### Diagnosing the Problem! ####

Indicatons that something's not right!

- **message**: A generic notification/diagnostic message produced by the message function;execution of the function continues
- **warning**: An indication that something is wrong but not necessarily fatal; execution of the function continues; generated by the **warning** function
- **error**: An indication that a fatal problem has occurred; execution stops; produced by the **stop** function
- **condition**: A generic concept for indicating that something unexpected can occur; programmerscan create their own conditions.

```{r}
log(-1)
```
```
printmessage <- function(x) {
 if (x > 0)
 print("x is greater than zero") else print("x is less than or equal to zero")
 invisible(x)
}
printmessage(1)
## [1] "x is greater than zero"
printmessage(NA)
## Error: missing value where TRUE/FALSE needed
```

```{r}
printmessage2 <- function(x) {
 if(is.na(x))
 print("x is a missing value!")
 else if(x > 0)
 print("x is greater than zero")
 else
 print("x is less than or equal to zero")
 invisible(x)
}

printmessage2(x)
```
How do you know that something is wrong with your function?

- What was your input? How did you call the function?
- What were you expecting? Output, messages, other results?
- What did you get?
- How does what you get differ from what you were expecting?
- Were your expectations correct in the first place?
- Can you reproduce the problem (exactly)?

#### Basic Debugging Tools ####
The primary tools for debugging functions in R are

- **traceback**: prints out the function call stack after an error occurs; does nothing if there is no error.
- **debug**: flags a function for "debug" mode which allows you to step through execution of a function one line at a time.
- **browser**: suspends the execution of a function wherever it is called and puts the function in debug mode
- **trace**: allows you to insert debugging code into a function a specific places
- **recover**: allows you to modify the error behavior so that you can browse the function call stack

There are interactive tools specifically designed to allow you to pick through a function. There is also the more blunt technique of inserting principle.cat statements in the function.

#### Debugging ####
##### Summary #####
- There are three main indications of a problem/condition: **message**, **warning**, **error**, only an **error** is fatal
- When analyzing a function with a problem, make sure you can reproduce the problem, clearly state your expectations and how the output differs from your expectation
- Interactive debugging tools **traceback**, **debug**, **browser**, **trace**, and **recover** can be used to find problematic code in functions
- Debugging tools are not a substitute for thinking!
